---
title: "Distributed DESeq2 Test"
output: html_notebook
---

# can DESeq be run in parts?
We have 1.5 Tb of GTEx salmon quant.sf files from over 17,000 sample. DESeq runs
on a single machine. Can we partiion the sample and run them seperately then 
recombine the part? i.e. the combind parts should be identical to result
creating running every thing on one machine

## General approach
1. get the normalized counts 
  normalization sums across all gene rows, and all sample columns
2. partion the original count matrix, and normalized counts
3. run DESeq on each part 
4. combin the parts

To test we will create a simple example by hand

Notes accessor functions are of the form

counts(dds) # get
counts(dds) < -value # set

```{r include=FALSE}
# produce a lot of output we do not want to see
library("DESeq2")
```


## step 1 create sample data

```{r}
nGenes = 8 
numCtrlSamples = 2
numTreatmentSamples = 2
nSamples = numCtrlSamples + numTreatmentSamples
nClasses = 2

# create the control samples
c1 <- seq(from=10, to=80, by=10)
c2 <- c1 + 2

# create the treatment samples
t1 <- c1

# over express 1 first and 3rd gene
overExprGenesList <- c( 1, 3 )
t1[ overExprGenesList ] <- c1[ overExprGenesList ] * 2**3

t2 <- t1 + 2

countMatrix <- matrix( c(c1, c2, t1, t2), ncol=nSamples)
countMatrix
```


## Step 2, run through the vignette quick start example

```{r}

# create the column meta data
sampleType <- c( rep('control', numCtrlSamples), rep('treatment', numTreatmentSamples) )
sampleType <- factor( sampleType )

colData <- data.frame( sampleType )

# check levels. by default R choose alphabetic order
# we need to make sure the control level is first or use the contrast argument
# in DESeq package. You can also use factor relevel()
levels(colData$sampleType)
cat(colData$sampleType)
cat("\n")
# create a DESeqDataSetFromMatrix object
dds <- DESeqDataSetFromMatrix(countData = countMatrix,
                              colData = colData,
                              design= ~ + sampleType)
```

```{r}
counts(dds)
```

```{r}
design( dds )
```

## We can not run the DESeq(dds) with the default values. 

we get an error msg

"all gene-wise dispersion estimates are within 2 orders of magnitude
from the minimum value, and so the standard curve fitting techniques will not work."

The error msg include a work around

TODO: use rnbinom() to generate our test data

ref:

* [What is the R rnbinom negative binomial dispersion parameter?](https://stats.stackexchange.com/questions/243899/what-is-the-r-rnbinom-negative-binomial-dispersion-parameter)
* [NegBinomial {stats}](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/NegBinomial.html)
* [rnegbin {MASS}](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/rnegbin.html)

## work around step 1) Estimate the size factors for a DESeqDataSet

default type is "ratio". Ratio uses the standard median ratio method introduced in DESeq. The size factor is the median ratio of the sample over a "pseudosample": for each gene, the geometric mean of all samples. 

```{r}
dds <- estimateSizeFactors(dds)
sizeFactors( dds )

cat("\n now that we have size factors we can get normalized counts\n")
counts(dds, normalized=TRUE) # get()
``` 

## work around step 2) Estimate the dispersions for a DESeqDataSet work around

```{r}
dds <- estimateDispersionsGeneEst(dds)
print( mcols(dds) )

dispersions(dds) <- mcols(dds)$dispGeneEst # set
cat("\n gene dispersions\n") 
dispersions(dds)
```

## work around step 3) Negative Binomial GLM fitting and Wald statistics: nbinomWaldTest

```{r}
dds <- nbinomWaldTest(dds)
print( mcols(dds) )
```

get the resuls
```{r}
# get a list of result coefficients
resultsNames( dds ) 
```

## save our expected results
```{r}
# results takes a name argument. by default we sampleType treatment vs control 
# see the documentation there are lots of arguments
expectedDESeqResults <- results(dds)
expectedDESeqResults
```

------ 
***
****

# Can we parition out data and get the same results?



## step 1 create DESeqDataSet for each partition

```{r}
p1Matrix = countMatrix[1:4 ,  ]
p2Matrix = countMatrix[5:8 ,  ]
```

create the meta data. This will be the same for all partitions. Each row must
include all the sample. We will create the partition by subseting the genes

```{r}
# create the column meta data
# it is important that 
pSampleType <- c( rep('control', numCtrlSamples), rep('treatment', numTreatmentSamples) )
pSampleType <- factor( pSampleType )

pColData <- data.frame( pSampleType )

# check levels. by default R choose alphabetic order
# we need to make sure the control level is first or use the contrast argument
# in DESeq package. You can also use factor relevel()
levels(pColData$pSampleType)
cat(pColData$pSampleType)
```

```{r}
# p1SampleType <- c( rep('control', numCtrlSamples) )
# p1SampleType <- factor( p1SampleType )
# 
# p1ColData <- data.frame( p1SampleType )
p1dds <- DESeqDataSetFromMatrix(countData = p1Matrix,
                              colData = pColData,
                              design= ~ + pSampleType)

counts( p1dds )
```

```{r}
# p2SampleType <- c( rep('treatment', numTreatmentSamples) )
# p2SampleType <- factor( p2SampleType )
# 
# p2ColData <- data.frame( p2SampleType )
p2dds <- DESeqDataSetFromMatrix(countData = p2Matrix,
                              colData = pColData,
                              design= ~ + pSampleType)

counts( p2dds )
```

## step 2) set the size factor calculated using full count matrix

```{r}
sizeFactors( p1dds ) <- sizeFactors( dds )
sizeFactors( p2dds ) <- sizeFactors( dds )
sizeFactors( p2dds ) # get
```

## step 3) work around step Estimate the dispersions for a DESeqDataSet work around

TODO hand craft better data

```{r}
p1dds <- estimateDispersionsGeneEst(p1dds)
print( mcols(p1dds) )

dispersions(p1dds) <- mcols(p1dds)$dispGeneEst # set
cat("\n gene dispersions\n") 
dispersions(p1dds)
```

```{r}
p2dds <- estimateDispersionsGeneEst(p2dds)
print( mcols(p2dds) )
dispersions(p2dds) <- mcols(p2dds)$dispGeneEst # set
cat("\n gene dispersions\n") 
dispersions(p2dds) 
```

## work around step 4) Negative Binomial GLM fitting and Wald statistics: nbinomWaldTest

```{r}
p1dds <- nbinomWaldTest( p1dds )
print( mcols(p1dds) )
```

```{r}
p2dds <- nbinomWaldTest( p2dds )
print( mcols(p2dds) )
```

## step 4 calculate the results
```{r}
# get a list of result coefficients
resultsNames( p1dds ) 

# results takes a name argument. by default we sampleType treatment vs control 
# see the documentation there are lots of arguments
p1ExpectedDESeqResults <- results(p1dds)
p1ExpectedDESeqResults
```

```{r}
# get a list of result coefficients
resultsNames( p2dds ) 

# results takes a name argument. by default we sampleType treatment vs control 
# see the documentation there are lots of arguments
p2ExpectedDESeqResults <- results(p2dds)
p2ExpectedDESeqResults
```